---
title: "Baseline Preprocessing"
author: "WGD"
date: "`r Sys.Date()`"
output: html_document

---

```{r setup, include = FALSE}
# Load required packages
library(here)
library(readxl)
library(dplyr)
library(janitor)
library(ggplot2)

# Define paths to input data
Apath <- here("data", "QuestionnairesA.xlsx")
Bpath <- here("data", "QuestionnairesB.xlsx")
```

```{r}
# Read files into dataframes A and B
A <- read_excel(Apath)
B <- read_excel(Bpath)
```

#Merge the baseline datasets
```{r}
baseline <- merge(A, B, by = "participant-id", all = TRUE, suffixes = c(".A", ".B"))
```

#Run the 'clean_names' function in the janitor package, so that variable names with hyphens become variable names with underscores (e.g., 'participant-id' -> 'participant_id') and all text in variable names becomes lowercase.
```{r}
baseline1 <- clean_names(baseline)
```
#Filter the data to include all 95 participants who completed Part 2 (3-day EMA period) of the parent study.
```{r}
analytic_sample <- c(102, 104, 107, 112, 113, 115, 116, 124, 125, 126, 127, 128, 133, 141, 150, 151, 154, 159, 163, 164, 166, 169, 175, 177, 180, 186, 188,
192, 195, 196, 197, 198, 199, 202, 206, 209, 210, 213, 214, 216, 218, 221, 222, 223, 233, 238, 240, 246, 252, 257, 262, 268, 273, 274, 278, 281, 282, 286, 287, 290, 291, 295, 297, 300, 301, 304, 310, 314, 320, 321, 329, 332, 340, 341, 342, 344, 346, 349, 351, 356, 362, 363, 369, 370, 372, 381, 383, 393, 394, 399, 416, 423, 427, 428, 430)
baseline2 <- baseline1[baseline1$participant_id %in% analytic_sample,]
```
#Next section containts measure-specific renaming and computing of baseline measures.

#FTND  (Fagerstrom, 2011; Heatherton et. al, 1991)
```{r}
#Rename ftnd_4 'cigsperday'
baseline2 <- baseline2 %>%
  rename(cigsperday = ftnd_4)
#Recode an item of the FTND, 'ftnd_4', so that an overall FTND_sum score can be calculated.

baseline2 <- baseline2 %>%
  mutate(ftnd_4 = case_when(
    cigsperday <= 10 ~ 1,
    cigsperday > 10 & cigsperday <= 20 ~ 2,
    cigsperday > 20 & cigsperday <= 30 ~ 3,
    cigsperday > 30 ~ 4
  ))

#Create a FTND_sum score
baseline2 <- baseline2 %>%
  mutate(ftnd_sum = ftnd_1 + ftnd_2 + ftnd_3 + ftnd_4 + ftnd_5 + ftnd_6)
```


#Positive and Negative Affect Schedule - Trait Version (PANAS-SF; Thompson, 2007; Watson et, al. 1988)
```{r}
#Create a subscale score for the positive and negative affect scales from the Positive and Negative Affect Schedule.
#Create the positive affect subscale
baseline2 <- baseline2 %>%
  mutate(panas_pos = q64_pan_1 + q64_pan_3 + q64_pan_5 + q64_pan_9 + q64_pan_10 + q64_pan_12 + q64_pan_14 + q64_pan_16 + q64_pan_17 + q64_pan_19)

#Create the negative affect subscale
baseline2 <- baseline2 %>%
  mutate(panas_neg = q64_pan_2 + q64_pan_4 + q64_pan_6 + q64_pan_7 + q64_pan_8 + q64_pan_11 + q64_pan_13 + q64_pan_15 + q64_pan_18 + q64_pan_20)
```

#Snaith-Hamilton Pleasure Scale (SHAPS;Snaith et. al., 1995)
#Recode the SHAPS variables based on author guidelines and create a sum score.
```{r}
#Create a new series of variables for this measure so that all 1s and 2s become 1 and all 3s and 4s become 0s. The new variables have the suffix "_sim", which is an abbreviation for "simple", because the authors refer to this system as the "simple scoring system"
baseline2 <- baseline2 %>%
  mutate(across(starts_with("q14_sha_"), ~ifelse(. %in% c(1, 2), 1, 0), .names = "{.col}_sim"))
#Create a sum score using the "_sim" variables
baseline2 <- baseline2 %>%
  mutate(shaps_sim_sum = q14_sha_1_sim + q14_sha_2_sim + q14_sha_3_sim + q14_sha_4_sim + q14_sha_5_sim + q14_sha_6_sim + q14_sha_7_sim + q14_sha_8_sim + q14_sha_9_sim + q14_sha_10_sim + q14_sha_11_sim + q14_sha_12_sim + q14_sha_13_sim + q14_sha_14_sim)
```

#Reformatting
#Change column types: columns containing only numbers become "integer", columns containing only letters or numbers-and-letters become "character", except if they have "date" in the title, in which case they become "POSIXct"
```{r}
# Function to change column types based on data and column names
 change_column_types <- function(df) {
  for (col_name in colnames(df)) {
    col_data <- df[[col_name]]

    # Skip columns already numeric or POSIXct
    if (is.numeric(col_data) || inherits(col_data, "POSIXct")) next

    # Trim character values
    if (is.character(col_data)) {
      trimmed <- trimws(col_data)
      numeric_check <- suppressWarnings(!is.na(as.numeric(trimmed)))
      
      if (all(numeric_check | is.na(trimmed))) {
        df[[col_name]] <- as.integer(trimmed)
      } else {
        df[[col_name]] <- trimmed
      }
    }
  }

  # Convert date columns
  for (col_name in colnames(df)) {
    if (grepl("_date_", col_name)) {
      df[[col_name]] <- as.POSIXct(df[[col_name]], format = "%m/%d/%Y")
    }
  }

  return(df)
}
baseline2 <- change_column_types(baseline2)
```

#Fix a typo
```{r}
baseline2 <- baseline2 %>%
  rename(total_household_income = total_houselhold_income)
```

#Subset the dataframe to only include variables relevant to the planned analyses and save the data
```{r}
saveRDS(
  baseline2[, c(
    "participant_id",
    "race",
    "age",
    "gender",
    "total_household_income",
    "bank_balance",
    "ftnd_sum",
    "cigsperday",
    "shaps_sim_sum",
    "panas_pos",
    "panas_neg"
  )],
  here::here("output", "baseline2.rds")
)
```

#References
#Carver, C. S., & White, T. L. (1994). Behavioral inhibition, behavioral activation, and affective responses to impending reward and punishment: The BIS/BAS scales. Journal of Personality and Social Psychology, 67(2), 319–333.
#Fagerström, K. (2011). Determinants of tobacco use and renaming the FTND to the Fagerström Test for Cigarette Dependence. Nicotine & Tobacco Research, 14(1), 75–78
#Heatherton, T. F., Kozlowski, L. T., Frecker, R. C., & FAGERSTROM, K. O. (1991). The Fagerström test for nicotine dependence: a revision of the Fagerstrom Tolerance Questionnaire. British journal of addiction, 86(9), 1119-1127.
#Snaith, R. P., Hamilton, M., Morley, S., Humayan, A., Hargreaves, D., & Trigwell, P. (1995). A scale for the assessment of hedonic tone: The Snaith–Hamilton Pleasure Scale. The British Journal of Psychiatry, 167(1), 99–103.
#Thompson, E. R. (2007). Development and validation of an internationally reliable short-form of the positive and negative affect schedule (PANAS). Journal of cross-cultural psychology, 38(2), 227-242.
#Watson, D., Clark, L. A., & Tellegen, A. (1988). Development and validation of brief measures of positive and negative affect: The PANAS scales. Journal of Personality and Social Psychology, 54(6), 1063–1070.