---
title: "Aims 1 & 2: Modeling Craving, Activity Appraisal, and Mood During Ad-Libitum Smoking"
author:
  - name: Walter G. Dyer, wdyer@psu.edu, Penn State University
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      number_sections: false
      toc_float: true
      toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Overview

This analysis tests two complementary hypotheses about how cigarette craving is related to emotional experience during ad-libitum smoking (Part 2, baseline):

**Aim 1: Pleasantness related affect**
- Craving moderates the effect of pleasantness on mood
- High craving *weakens* the association between pleasant activities and positive affect
- Expected: negative interaction between pleasantness and craving

**Aim 2: Unpleasantness related affect**
- Craving moderates the effect of unpleasantness on mood
- High craving *strengthens* the association between unpleasant activities and negative affect
- Expected: positive interaction between unpleasantness and craving

Both analyses use multilevel mixed-effects regression with sequential model building to evaluate within-person associations, moderation effects, and individual differences in reactivity patterns.

**Note:** All analyses are conducted on Part 2 data only. For Aim 1, mood is filtered to > 50 (positive mood); for Aim 2, mood is filtered to < 50 (negative mood). Each aim is run on both all available data and valid-cases-only as a sensitivity check.

---

# Load packages and data

```{r packages, message=FALSE, warning=FALSE}
# Data management
library(dplyr)
library(tidyr)
library(here)

# Multilevel modeling
library(lme4)
library(lmerTest)

# Model diagnostics and summaries
library(performance)
library(sjPlot)
library(broom.mixed)

# Visualization
library(ggplot2)

# Tables
library(knitr)
library(kableExtra)
```

```{r load-data}
# Load preprocessed dataset
study_sample <- readRDS(here("output", "cleaned_study_sample.RDS"))

# Load validity function
check_validity <- function(df) {
  mood_ok          <- !is.na(df$mood)
  before_survey_ok <- !is.na(df$before_survey)
  smoke_ok         <- !is.na(df$smoke)
  branch_ok        <- !is.na(df$pleasant) | !is.na(df$unpleasant)
  as.integer(mood_ok & before_survey_ok & smoke_ok & branch_ok)
}
```

---

# AIM 1: Pleasantness Related Affect and Craving

### Model 1a: Unconditional

```{r aim1-all-1a}
aim1_1a_all <- lmer(
  mood ~ 1 + (1 | participant_id),
  data = aim1_all_available,
  REML = TRUE
)

summary(aim1_1a_all)
```

### Model 1b: Main effects

```{r aim1-all-1b}
aim1_1b_all <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pm + smoke_pm + 
    (1 | participant_id),
  data = aim1_all_available,
  REML = TRUE
)

summary(aim1_1b_all)

anova(aim1_1a_all, aim1_1b_all)
```

### Model 1c: Add interaction (fixed)

```{r aim1-all-1c}
aim1_1c_all <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc + 
    pleasant_pm + smoke_pm + 
    (1 | participant_id),
  data = aim1_all_available,
  REML = TRUE
)

summary(aim1_1c_all)

anova(aim1_1b_all, aim1_1c_all)
```

### Model 1d: Random slope for pleasantness

```{r aim1-all-1d}
aim1_1d_all <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc + 
    pleasant_pm + smoke_pm + 
    (1 + pleasant_pmc || participant_id),
  data = aim1_all_available,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim1_1d_all)

anova(aim1_1c_all, aim1_1d_all)
```

### Model 1e: Random slope for craving

```{r aim1-all-1e}
aim1_1e_all <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc + 
    pleasant_pm + smoke_pm + 
    (1 + pleasant_pmc + smoke_pmc || participant_id),
  data = aim1_all_available,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim1_1e_all)

anova(aim1_1d_all, aim1_1e_all)
```

### Model 1f: Random slope for interaction

```{r aim1-all-1f}
aim1_1f_all <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc + 
    pleasant_pm + smoke_pm + 
    (1 + pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc || participant_id),
  data = aim1_all_available,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim1_1f_all)

anova(aim1_1e_all, aim1_1f_all)
```

### Model comparison summary (Aim 1, All Available)

```{r aim1-all-summary}
aim1_all_models <- list(
  "1a: Unconditional" = aim1_1a_all,
  "1b: Main effects" = aim1_1b_all,
  "1c: + Interaction (fixed)" = aim1_1c_all,
  "1d: + Random slope (pleasant)" = aim1_1d_all,
  "1e: + Random slope (smoke)" = aim1_1e_all,
  "1f: + Random slope (interaction)" = aim1_1f_all
)

kable(
  data.frame(
    Model = names(aim1_all_models),
    AIC = sapply(aim1_all_models, AIC),
    BIC = sapply(aim1_all_models, BIC),
    Deviance = sapply(aim1_all_models, deviance),
    df = sapply(aim1_all_models, function(x) attr(logLik(x), "df"))
  ),
  caption = "Aim 1 Model Fit Comparison (All Available Data)",
  digits = 2
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

## Aim 1: Valid Cases Only

### Model 1a: Unconditional

```{r aim1-valid-1a}
aim1_1a_valid <- lmer(
  mood ~ 1 + (1 | participant_id),
  data = aim1_valid_only,
  REML = TRUE
)

summary(aim1_1a_valid)
```

### Model 1b: Main effects

```{r aim1-valid-1b}
aim1_1b_valid <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pm + smoke_pm + 
    (1 | participant_id),
  data = aim1_valid_only,
  REML = TRUE
)

summary(aim1_1b_valid)

anova(aim1_1a_valid, aim1_1b_valid)
```

### Model 1c: Add interaction (fixed)

```{r aim1-valid-1c}
aim1_1c_valid <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc + 
    pleasant_pm + smoke_pm + 
    (1 | participant_id),
  data = aim1_valid_only,
  REML = TRUE
)

summary(aim1_1c_valid)

anova(aim1_1b_valid, aim1_1c_valid)
```

### Model 1d: Random slope for pleasantness

```{r aim1-valid-1d}
aim1_1d_valid <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc + 
    pleasant_pm + smoke_pm + 
    (1 + pleasant_pmc || participant_id),
  data = aim1_valid_only,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim1_1d_valid)

anova(aim1_1c_valid, aim1_1d_valid)
```

### Model 1e: Random slope for craving

```{r aim1-valid-1e}
aim1_1e_valid <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc + 
    pleasant_pm + smoke_pm + 
    (1 + pleasant_pmc + smoke_pmc || participant_id),
  data = aim1_valid_only,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim1_1e_valid)

anova(aim1_1d_valid, aim1_1e_valid)
```

### Model 1f: Random slope for interaction

```{r aim1-valid-1f}
aim1_1f_valid <- lmer(
  mood ~ pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc + 
    pleasant_pm + smoke_pm + 
    (1 + pleasant_pmc + smoke_pmc + pleasant_pmc:smoke_pmc || participant_id),
  data = aim1_valid_only,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim1_1f_valid)

anova(aim1_1e_valid, aim1_1f_valid)
```

### Model comparison summary (Aim 1, Valid Cases Only)

```{r aim1-valid-summary}
aim1_valid_models <- list(
  "1a: Unconditional" = aim1_1a_valid,
  "1b: Main effects" = aim1_1b_valid,
  "1c: + Interaction (fixed)" = aim1_1c_valid,
  "1d: + Random slope (pleasant)" = aim1_1d_valid,
  "1e: + Random slope (smoke)" = aim1_1e_valid,
  "1f: + Random slope (interaction)" = aim1_1f_valid
)

kable(
  data.frame(
    Model = names(aim1_valid_models),
    AIC = sapply(aim1_valid_models, AIC),
    BIC = sapply(aim1_valid_models, BIC),
    Deviance = sapply(aim1_valid_models, deviance),
    df = sapply(aim1_valid_models, function(x) attr(logLik(x), "df"))
  ),
  caption = "Aim 1 Model Fit Comparison (Valid Cases Only)",
  digits = 2
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# AIM 2: Stress Reactivity

## Aim 2: All Available Data

### Model 2a: Unconditional

```{r aim2-all-2a}
aim2_2a_all <- lmer(
  mood ~ 1 + (1 | participant_id),
  data = aim2_all_available,
  REML = TRUE
)

summary(aim2_2a_all)
```

### Model 2b: Main effects

```{r aim2-all-2b}
aim2_2b_all <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pm + smoke_pm + 
    (1 | participant_id),
  data = aim2_all_available,
  REML = TRUE
)

summary(aim2_2b_all)

anova(aim2_2a_all, aim2_2b_all)
```

### Model 2c: Add interaction (fixed)

```{r aim2-all-2c}
aim2_2c_all <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc + 
    unpleasant_pm + smoke_pm + 
    (1 | participant_id),
  data = aim2_all_available,
  REML = TRUE
)

summary(aim2_2c_all)

anova(aim2_2b_all, aim2_2c_all)
```

### Model 2d: Random slope for unpleasantness

```{r aim2-all-2d}
aim2_2d_all <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc + 
    unpleasant_pm + smoke_pm + 
    (1 + unpleasant_pmc || participant_id),
  data = aim2_all_available,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim2_2d_all)

anova(aim2_2c_all, aim2_2d_all)
```

### Model 2e: Random slope for craving

```{r aim2-all-2e}
aim2_2e_all <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc + 
    unpleasant_pm + smoke_pm + 
    (1 + unpleasant_pmc + smoke_pmc || participant_id),
  data = aim2_all_available,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim2_2e_all)

anova(aim2_2d_all, aim2_2e_all)
```

### Model 2f: Random slope for interaction

```{r aim2-all-2f}
aim2_2f_all <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc + 
    unpleasant_pm + smoke_pm + 
    (1 + unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc || participant_id),
  data = aim2_all_available,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim2_2f_all)

anova(aim2_2e_all, aim2_2f_all)
```

### Model comparison summary (Aim 2, All Available)

```{r aim2-all-summary}
aim2_all_models <- list(
  "2a: Unconditional" = aim2_2a_all,
  "2b: Main effects" = aim2_2b_all,
  "2c: + Interaction (fixed)" = aim2_2c_all,
  "2d: + Random slope (unpleasant)" = aim2_2d_all,
  "2e: + Random slope (smoke)" = aim2_2e_all,
  "2f: + Random slope (interaction)" = aim2_2f_all
)

kable(
  data.frame(
    Model = names(aim2_all_models),
    AIC = sapply(aim2_all_models, AIC),
    BIC = sapply(aim2_all_models, BIC),
    Deviance = sapply(aim2_all_models, deviance),
    df = sapply(aim2_all_models, function(x) attr(logLik(x), "df"))
  ),
  caption = "Aim 2 Model Fit Comparison (All Available Data)",
  digits = 2
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

## Aim 2: Valid Cases Only

### Model 2a: Unconditional

```{r aim2-valid-2a}
aim2_2a_valid <- lmer(
  mood ~ 1 + (1 | participant_id),
  data = aim2_valid_only,
  REML = TRUE
)

summary(aim2_2a_valid)
```

### Model 2b: Main effects

```{r aim2-valid-2b}
aim2_2b_valid <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pm + smoke_pm + 
    (1 | participant_id),
  data = aim2_valid_only,
  REML = TRUE
)

summary(aim2_2b_valid)

anova(aim2_2a_valid, aim2_2b_valid)
```

### Model 2c: Add interaction (fixed)

```{r aim2-valid-2c}
aim2_2c_valid <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc + 
    unpleasant_pm + smoke_pm + 
    (1 | participant_id),
  data = aim2_valid_only,
  REML = TRUE
)

summary(aim2_2c_valid)

anova(aim2_2b_valid, aim2_2c_valid)
```

### Model 2d: Random slope for unpleasantness

```{r aim2-valid-2d}
aim2_2d_valid <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc + 
    unpleasant_pm + smoke_pm + 
    (1 + unpleasant_pmc || participant_id),
  data = aim2_valid_only,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim2_2d_valid)

anova(aim2_2c_valid, aim2_2d_valid)
```

### Model 2e: Random slope for craving

```{r aim2-valid-2e}
aim2_2e_valid <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc + 
    unpleasant_pm + smoke_pm + 
    (1 + unpleasant_pmc + smoke_pmc || participant_id),
  data = aim2_valid_only,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim2_2e_valid)

anova(aim2_2d_valid, aim2_2e_valid)
```

### Model 2f: Random slope for interaction

```{r aim2-valid-2f}
aim2_2f_valid <- lmer(
  mood ~ unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc + 
    unpleasant_pm + smoke_pm + 
    (1 + unpleasant_pmc + smoke_pmc + unpleasant_pmc:smoke_pmc || participant_id),
  data = aim2_valid_only,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(aim2_2f_valid)

anova(aim2_2e_valid, aim2_2f_valid)
```

### Model comparison summary (Aim 2, Valid Cases Only)

```{r aim2-valid-summary}
aim2_valid_models <- list(
  "2a: Unconditional" = aim2_2a_valid,
  "2b: Main effects" = aim2_2b_valid,
  "2c: + Interaction (fixed)" = aim2_2c_valid,
  "2d: + Random slope (unpleasant)" = aim2_2d_valid,
  "2e: + Random slope (smoke)" = aim2_2e_valid,
  "2f: + Random slope (interaction)" = aim2_2f_valid
)

kable(
  data.frame(
    Model = names(aim2_valid_models),
    AIC = sapply(aim2_valid_models, AIC),
    BIC = sapply(aim2_valid_models, BIC),
    Deviance = sapply(aim2_valid_models, deviance),
    df = sapply(aim2_valid_models, function(x) attr(logLik(x), "df"))
  ),
  caption = "Aim 2 Model Fit Comparison (Valid Cases Only)",
  digits = 2
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Interpretation Checkpoints

## Key questions to guide your sequential decision-making:

**Aim 1: Reward Reactivity**
- Does the pleasantness × craving interaction (Model 1c) improve fit and is it significant in the expected direction (negative)?
- Does allowing random slopes for pleasantness and/or craving meaningfully improve the model?
- How much variability in the interaction slope exists across individuals?
- Do results replicate in the valid-cases-only sensitivity analysis?

**Aim 2: Stress Reactivity**
- Does the unpleasantness × craving interaction (Model 2c) improve fit and is it significant in the expected direction (positive)?
- Does allowing random slopes for unpleasantness and/or craving meaningfully improve the model?
- How much variability in the interaction slope exists across individuals?
- Do results replicate in the valid-cases-only sensitivity analysis?

**General Considerations**
- Which models converge successfully? Are there optimizer warnings?
- Do AIC/BIC comparisons and nested model tests align in their model preferences?
- How sensitive are findings to the all-available vs. valid-only distinction?
- Are effect sizes meaningful in addition to statistical significance?

---

# References

Dyer, W.G. (2024). Dissertation analyses leveraging the DVAL project EMA data.
