---
title: "Data characteristics report"
author:
  - name: Walter G. Dyer, wdyer@psu.edu, Penn State University
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      number_sections: false
      toc_float: true
      toc_depth: 2
---


<!-- Javascript code to zoom in and out in the plots. -->
<!-- Code from Radovan Miletić: https://stackoverflow.com/questions/56361986/zoom-function-in-rmarkdown-html-plot -->
<script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"></script>
<script type="text/javascript">
$(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
    $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
    $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
    $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
});
</script>

<!--------- INSTRUCTIONS ----------
This template would help you to structure and share the characteristic and the characteristics of your preprocessed dataset. 
It is based on the ESM preprocessing framework described in Revol and al. (under review). 
In addition, it is associated with an R packages, the esmtools package (package.esmtools.com), 
and a website to provide tutorials and R code for each step: preprocess.esmtools.com. 
Please, follow the instructions commented and add descriptions above each chunk you create. 
Note that the commented instructions are not displayed in the rendered document. 
<!--------------------------------->



<!-- This first chunk define the global settings and hidden variables of the file. -->
```{r, include=FALSE}
# For scroll bar when the output is longer than what is defined in the chunk:   {r, max.height='600px'}
options(width = 10000)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})

knitr::opts_chunk$set(message=FALSE, warning=FALSE) # set multiple global options here
```
# Import
```{r}
#Packages
library(here)
library(dplyr)
library(tidyr)
library(tableone)
```

```{r}
# Read in data
study_sample <- readRDS(here("output", "cleaned_study_sample.RDS"))
```

#1.	What are the demographic characteristics of the sample?

```{r}
#"Table 1."
library(tableone)
library(dplyr)
library(gt)

# Recode race and gender as factors with labels
study_sample <- study_sample %>%
  mutate(
    race = factor(
      race,
      levels = 1:7,
      labels = c(
        "American Indian or Alaska Native",
        "Asian",
        "Black or African-American",
        "Native Hawaiian or Other Pacific Islander",
        "White",
        "More than one race",
        "Unknown or prefer not to answer"
      )
    ),
    gender = factor(
      gender,
      levels = 1:2,
      labels = c("Male", "Female")
    )
  )

# Variables for Table 1
vars <- c(
  "race",
  "age",
  "gender",
  "total_household_income",
  "bank_balance",
  "ftnd_sum",
  "cigsperday",
  "shaps_sim_sum",
  "panas_pos",
  "panas_neg"
)

catVars <- c("race", "gender")

# Create Table 1
table1 <- CreateTableOne(
  vars = vars,
  data = study_sample,
  factorVars = catVars
)

# Convert Table 1 to a data.frame for gt
table1_df <- print(table1, showAllLevels = TRUE, quote = FALSE, printToggle = FALSE) %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Variable") %>%
  rename(Statistic = `Overall`)

# Optional: remove % sign for cleaner look if desired
table1_df$Statistic <- gsub("%", "", table1_df$Statistic)

# Make GT table
table1_df %>%
  gt() %>%
  tab_header(
    title = "Table 1. Participant Characteristics"
  ) %>%
  cols_label(
    Variable = "Variable",
    Statistic = "Overall"
  ) %>%
  fmt_number(
    columns = vars[vars != "race" & vars != "gender"], # format numeric variables
    decimals = 2
  ) %>%
  opt_table_font(
    font = list(
      default_fonts()
    )
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(16),
    data_row.padding = px(5)
  )
```






#2.	What are the intraclass correlations (ICCs) for key EMA variables (e.g., craving, unpleasantness, pleasantness, and mood)?
  #a.	This will help distinguish between- and within-person variance and justify modeling decisions
#3.	Means for mood and craving across unpleasant, pleasant, and neutral event appraisal conditions.
#4.	What is the effect of the abstinence incentive?
#5.	Person-level cigarette use behavior over the course of the study.
#6.	Number of available cases for each Aim.
  #a.	This will involve the creation of a consort diagram
#7.	Descriptives of compliance. Are there significant trends in non-compliance that should be accounted for (e.g., participant characteristics)?
#8.	Do the CO data align with the EMA data on cigarette smoking for each participant?
