---
title: "EMA Data Characteristics Report"
author:
  - name: Walter G. Dyer, wdyer@psu.edu, Penn State University
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      number_sections: false
      toc_float: true
      toc_depth: 3
---

<!-- Javascript code to zoom in and out in the plots. -->
<!-- Code from Radovan Miletić: https://stackoverflow.com/questions/56361986/zoom-function-in-rmarkdown-html-plot -->
<script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"></script>
<script type="text/javascript">
$(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    $('img:not(.zoomImg)').click(function() {
    $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
    $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    $('img.zoomImg').click(function() {
    $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
});
</script>

<style>
/* esmtools default label styles */
.esm-issue{
}

/* Committee-specific callout */
.committee-note {
  background-color: #f0f4ff;
  border-left: 4px solid #4a6fa5;
  border-radius: 4px;
  padding: 8px 12px;
  margin: 8px 0;
  font-size: 0.9em;
  color: #2c3e6b;
}

.committee-note .committee-label {
  font-weight: bold;
  text-transform: uppercase;
  letter-spacing: 0.05em;
  font-size: 0.8em;
  color: #4a6fa5;
  display: block;
  margin-bottom: 2px;
}
/* Section summary callout */
.section-summary {
  background-color: #f8f8f8;
  border-left: 4px solid #aaaaaa;
  border-radius: 4px;
  padding: 8px 12px;
  margin: 16px 0 8px 0;
  font-size: 0.9em;
  color: #333333;
}

.section-summary .summary-label {
  font-weight: bold;
  text-transform: uppercase;
  letter-spacing: 0.05em;
  font-size: 0.8em;
  color: #888888;
  display: block;
  margin-bottom: 2px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(esmtools)
```

`r button(text = "Supplementary")`

```{r packages, message=FALSE, warning=FALSE}
# Data management
library(dplyr)
library(tidyr)
library(here)
library(htmltools)
library(purrr)

# EMA preprocessing tools
library(esmtools)

# Descriptive statistics
library(skimr)

# Multilevel modeling (for ICC estimation)
library(lme4)

# Missing values
library(naniar)
library(visdat)

# Plotting
library(ggplot2)

# Dates and times
library(lubridate)
library(hms)

# Tables
library(knitr)
library(kableExtra)
```

```{r functions}
# Section summary function
section_summary <- function(text) {
  htmltools::HTML(paste0(
    '<div class="section-summary">',
    '<span class="summary-label">Summary</span>',
    text,
    '</div>'
  ))
}

# Committee callout function
committee_note <- function(text) {
  htmltools::HTML(paste0(
    '<div class="committee-note">',
    '<span class="committee-label">Committee</span>',
    text,
    '</div>'
  ))
}

# Validity function (carried over from preprocessing report)
check_validity <- function(df) {
  mood_ok          <- !is.na(df$mood)
  before_survey_ok <- !is.na(df$before_survey)
  smoke_ok         <- !is.na(df$smoke)
  branch_ok        <- !is.na(df$pleasant) | !is.na(df$unpleasant)
  as.integer(mood_ok & before_survey_ok & smoke_ok & branch_ok)
}
```

```{r load-data}
# Load preprocessed dataset from Report 1 output
study_sample <- readRDS(here("output", "cleaned_study_sample.RDS"))

# Part 6 completer IDs (any participant with Part 6 data)
part6_completers <- study_sample %>%
  filter(study_part == 6) %>%
  distinct(participant_id) %>%
  pull(participant_id)

# Valid Part 6 completer IDs (excludes participants with zero valid Part 6 obs,
# e.g., participant 223 who had missing branch variable data throughout Part 6)
valid_completers <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1, study_part == 6) %>%
  distinct(participant_id) %>%
  pull(participant_id)
```

`r endbutton()`

---

# Overview

This report characterizes the preprocessed EMA dataset for Walter Dyer's dissertation from the DVAL project parent study following Revol et al.'s (2024) data characteristics report guidelines. It describes the quality of the data collection, the properties of the key EMA variables, and selected preliminary analyses. Content flagged with a **Committee** label addresses questions raised during the proposal defense.

The preprocessed dataset was produced by the companion EMA Preprocessing Report and is loaded directly from that report's output. All analyses here are conducted on the fully preprocessed dataset.

**Study summary:** Participants completed EMA surveys 6 times per day within predetermined 2-hour intervals. Part 2 (Days 1--3) was a baseline period during which participants were instructed to smoke at their normal rate. Part 6 (Days 4--10) was an incentivized abstinence period during which participants could earn daily cash bonuses for verified cigarette abstinence. EMA, baseline, and biochemical verification (CO) data are included in the preprocessed dataset.

---

# Variable Descriptions

This section provides an overview of all variables in the preprocessed dataset, including their type, role, and expected range.

`r button(text = "Description")`

```{r variable-descriptions}
# Overview of variable names and types
glimpse(study_sample)
```

```{r variable-table}
# Summary table of key variables
var_info <- tibble::tribble(
  ~Variable,                  ~Type,          ~Role,                        ~Range_or_Values,
  "participant_id",           "Numeric",       "Identifier",                 "Unique per participant",
  "session_id",               "Numeric",       "Identifier",                 "Unique per beep",
  "study_part",               "Numeric",       "Study design",               "2 = baseline, 6 = incentive",
  "init_time",                "POSIXct",       "Timestamp",                  "Datetime",
  "completion_time",          "POSIXct",       "Timestamp",                  "Datetime",
  "obsno",                    "Numeric",       "Time variable",              "1-60",
  "dayno",                    "Numeric",       "Time variable",              "1-10",
  "beepno",                   "Numeric",       "Time variable",              "1-6",
  "duration",                 "Numeric",       "Time-invariant",             "Days of participation",
  "mood",                     "Numeric",       "EMA outcome",                "1-100",
  "smoke",                    "Numeric",       "EMA outcome",                "1-100",
  "before_survey",            "Categorical",   "EMA branching item",         "Pleasant/Unpleasant/Neutral",
  "pleasant",                 "Numeric",       "EMA outcome (branch)",       "1-100",
  "unpleasant",               "Numeric",       "EMA outcome (branch)",       "1-100",
  "prior_mood",               "Numeric",       "EMA outcome",                "1-100",
  "smoked_since_last_survey", "Numeric",       "EMA smoking behavior",       "0-6",
  "age",                      "Numeric",       "Baseline covariate",         "Years",
  "gender",                   "Categorical",   "Baseline covariate",         "Categorical",
  "race",                     "Categorical",   "Baseline covariate",         "Categorical",
  "total_household_income",   "Numeric",       "Baseline covariate",         "Dollars per year",
  "bank_balance",             "Numeric",       "Baseline covariate",         "Dollars",
  "ftnd_sum",                 "Numeric",       "Baseline covariate",         "0-10",
  "cigsperday",               "Numeric",       "Baseline covariate",         "Cigarettes smoked per day",
  "shaps_sim_sum",            "Numeric",       "Baseline covariate",         "0-14",
  "panas_pos",                "Numeric",       "Baseline covariate",         "10-50",
  "panas_neg",                "Numeric",       "Baseline covariate",         "10-50",
  "co_level",                 "Numeric",       "Biochemical verification",   "parts per million",
  "co_date",                  "Date",          "Biochemical verification",   "Date",
  "co_state_smoked",          "Logical",       "Biochemical verification",   "TRUE/FALSE",
  "co_correct_procedure",     "Logical",       "Biochemical verification",   "TRUE/FALSE",
  "q23",                      "Numeric",       "Baseline covariate",         "Years",
  "q24",                      "Numeric",       "Baseline covariate",         "Years",
  "q25",                      "Numeric",       "Baseline covariate",         "Years",
  "q71",                      "Numeric",       "Baseline covariate",         "Times in past 30 days",
  "q80",                      "Numeric",       "Baseline covariate",         "Times in past 30 days"
)

kable(var_info, caption = "Variable descriptions for the preprocessed DVAL EMA dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

`r endbutton()`

---

# Sample Description

`r committee_note("This section addresses committee questions about the demographic characteristics of the sample and the flow of participants through the study.")`

## Demographics

`r committee_note("The committee requested a description of the demographic characteristics of the sample.")`

`r button(text = "Description")`

```{r demographics}
# Pull one row per participant (time-invariant variables)
baseline_chars <- study_sample %>%
  group_by(participant_id) %>%
  slice(1) %>%
  ungroup()

cat("Total participants:", nrow(baseline_chars), "\n")

# --- Table 1: Continuous variables ---
cont_vars <- tibble::tribble(
  ~Variable,                          ~col,
  "Age (years)",                      "age",
  "Cigarettes per day (baseline)",    "cigsperday",
  "FTND total score",                 "ftnd_sum",
  "SHAPS score",                      "shaps_sim_sum",
  "PANAS positive affect",            "panas_pos",
  "PANAS negative affect",            "panas_neg",
  "Bank balance ($)",                 "bank_balance",
  "Age at first puff",                "q23",
  "Age at first full cigarette",      "q24",
  "Age at regular smoking onset",     "q25",
  "E-cigarette use (past 30 days)",   "q71",
  "Cannabis use (past 30 days)",      "q80"
)

cont_table <- cont_vars %>%
  rowwise() %>%
  mutate(
    n    = sum(!is.na(baseline_chars[[col]])),
    Mean = round(mean(baseline_chars[[col]], na.rm = TRUE), 2),
    SD   = round(sd(baseline_chars[[col]], na.rm = TRUE), 2),
    Min  = round(min(baseline_chars[[col]], na.rm = TRUE), 2),
    Max  = round(max(baseline_chars[[col]], na.rm = TRUE), 2)
  ) %>%
  ungroup() %>%
  select(Variable, n, Mean, SD, Min, Max)

cont_table %>%
  kable(caption = "Descriptive statistics for continuous baseline variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# --- Table 2: Gender ---
cat("\n")
baseline_chars %>%
  mutate(gender = as.character(gender)) %>%
  count(gender) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  rename(Gender = gender, N = n, `%` = pct) %>%
  kable(caption = "Gender") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# --- Table 3: Race ---
race_labels <- c(
  "1" = "American Indian or Alaska Native",
  "2" = "Asian",
  "3" = "Black or African-American",
  "4" = "Native Hawaiian or Other Pacific Islander",
  "5" = "White",
  "6" = "More than one race",
  "7" = "Unknown or prefer not to answer"
)

baseline_chars %>%
  mutate(race = recode(as.character(race), !!!race_labels)) %>%
  count(race) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  rename(Race = race, N = n, `%` = pct) %>%
  kable(caption = "Race/ethnicity") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# --- Table 4: Household income ---
baseline_chars %>%
  mutate(total_household_income = as.character(total_household_income)) %>%
  count(total_household_income) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  arrange(total_household_income) %>%
  rename(`Household Income` = total_household_income, N = n, `%` = pct) %>%
  kable(caption = "Household income — all categories shown") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

`r endbutton()`

## CONSORT Diagram

`r committee_note("The committee requested a CONSORT diagram to document participant flow through the study.")`

```{r consort, fig.width=8, fig.height=12}
library(DiagrammeR)

# Dynamically computed from study_sample
n_part2 <- study_sample %>%
  filter(study_part == 2) %>%
  distinct(participant_id) %>%
  nrow()

n_part6 <- study_sample %>%
  filter(study_part == 6) %>%
  distinct(participant_id) %>%
  nrow()

n_dropout <- n_part2 - n_part6

grViz(paste0('
digraph consort {

  graph [layout = dot, rankdir = TB, fontname = "Helvetica"]
  node  [shape = rectangle, fontname = "Helvetica", fontsize = 11,
         style = filled, fillcolor = white, margin = "0.2,0.1"]
  edge  [fontname = "Helvetica", fontsize = 10]

  # Main flow nodes
  screened   [label = "Screened via telephone\n(n = 1,187)"]
  consented  [label = "Consented\n(n = 139)"]
  baseline   [label = "Completed baseline laboratory visit\n(n = 118)"]
  part2      [label = "Completed 3-day baseline EMA and\nmet criteria for continuation\n(n = ', n_part2, ')"]
  mock       [label = "Completed mock scanner training visit\n(n = 82)"]
  randomized [label = "Randomized to scan order A (n = 41)\nor scan order B (n = 41)"]
  part6      [label = "Completed 7-day EMA period\nwith incentivized abstinence\n(n = ', n_part6, ')"]

  # Exclusion nodes
  excl1 [label = "Declined to participate (n = 189)\nNot eligible (n = 858)\nUnable to schedule: COVID-19 shutdown (n = 1)",
          shape = rectangle, fillcolor = "#f9f9f9"]
  excl2 [label = "Declined to continue (n = 2)\nScreened out: CPD below cutoff (n = 3)\nScreened out: baseline CO below cutoff (n = 5)\nScreened out: MRI-related exclusion (n = 6)\nScreened out: substance dependence on MINI (n = 5)",
          shape = rectangle, fillcolor = "#f9f9f9"]
  excl3 [label = "Declined to continue (n = 11)\nUnable to continue: COVID-19 shutdown (n = 3)\nWithdrawn: protocol compliance below threshold (n = 9)",
          shape = rectangle, fillcolor = "#f9f9f9"]
  excl4 [label = "Withdrawn: claustrophobia/discomfort (n = 6)\nWithdrawn: experimental CO above cutoff (n = 4)\nWithdrawn: MRI-related exclusion (n = 2)\nWithdrawn: repeated cancellations (n = 1)",
          shape = rectangle, fillcolor = "#f9f9f9"]
  excl5 [label = "Declined to continue (n = 10)\nDied (n = 1)",
          shape = rectangle, fillcolor = "#f9f9f9"]
  excl6 [label = "Dropout between Part 2 and Part 6\n(n = ', n_dropout, ')",
          shape = rectangle, fillcolor = "#f9f9f9"]

  # Main flow edges
  screened   -> consented
  consented  -> baseline
  baseline   -> part2
  part2      -> mock
  mock       -> randomized
  randomized -> part6

  # Exclusion edges
  screened   -> excl1   [arrowhead = none, style = dashed]
  consented  -> excl2   [arrowhead = none, style = dashed]
  baseline   -> excl3   [arrowhead = none, style = dashed]
  part2      -> excl4   [arrowhead = none, style = dashed]
  randomized -> excl5   [arrowhead = none, style = dashed]
  part6      -> excl6   [arrowhead = none, style = dashed]
}
'))
```

---

`r section_summary("The analytic sample comprises participants from the DVAL parent study who completed the 3-day baseline EMA period (Part 2). A subset of these participants also completed the 7-day incentivized abstinence EMA period (Part 6). Demographic characteristics and participant flow are documented above. See the CONSORT diagram for a full accounting of enrollment, attrition, and study completion.")`

# Response Behavior Quality

This section examines how well participants engaged with the EMA protocol.

## Compliance

```{r compliance-summary}
compliance_df <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  group_by(participant_id) %>%
  summarise(
    valid_beeps     = sum(valid, na.rm = TRUE),
    max_dayno       = max(dayno, na.rm = TRUE),
    total_possible  = if_else(max_dayno <= 3, 18L, 60L),
    completion_group = if_else(max_dayno <= 3,
                               "Part 2 Only (3 days)",
                               "Parts 2 + 6 (10 days)"),
    compliance      = valid_beeps / total_possible
  ) %>%
  ungroup()

# Summary by group
compliance_df %>%
  group_by(completion_group) %>%
  summarise(
    n                = n(),
    mean_valid_beeps = round(mean(valid_beeps), 1),
    mean_possible    = round(mean(total_possible), 1),
    mean_compliance  = round(mean(compliance), 3),
    min_compliance   = round(min(compliance), 3),
    max_compliance   = round(max(compliance), 3)
  ) %>%
  kable(caption = "Compliance summary by study completion group") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r compliance-plot}
ggplot(compliance_df, aes(x = compliance, fill = completion_group)) +
  geom_histogram(binwidth = 0.05, boundary = 0, color = "white") +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  facet_wrap(~ completion_group, ncol = 1) +
  labs(
    title = "Distribution of participant compliance by study completion group",
    x     = "Compliance (% of possible EMA prompts)",
    y     = "Number of participants"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Compliance Trends and Participant Characteristics

`r committee_note("The committee asked whether there are significant trends in non-compliance that should be accounted for (e.g., participant characteristics).")`

```{r compliance-trends}
# Merge compliance with baseline characteristics
compliance_chars <- compliance_df %>%
  left_join(
    study_sample %>%
      group_by(participant_id) %>%
      slice(1) %>%
      ungroup() %>%
      select(participant_id, age, gender, race,
             ftnd_sum, cigsperday, panas_pos, panas_neg),
    by = "participant_id"
  )

# Correlations between compliance and continuous baseline variables,
# with significance stars
cor_vars <- c("compliance", "age", "ftnd_sum", "cigsperday", "panas_pos", "panas_neg")

cor_data <- compliance_chars %>%
  select(all_of(cor_vars))

n_vars <- length(cor_vars)
r_mat  <- matrix(NA, n_vars, n_vars, dimnames = list(cor_vars, cor_vars))
p_mat  <- matrix(NA, n_vars, n_vars, dimnames = list(cor_vars, cor_vars))

for (i in seq_len(n_vars)) {
  for (j in seq_len(n_vars)) {
    if (i == j) {
      r_mat[i, j] <- 1
      p_mat[i, j] <- NA
    } else {
      test         <- cor.test(cor_data[[i]], cor_data[[j]],
                               use = "pairwise.complete.obs")
      r_mat[i, j] <- round(test$estimate, 3)
      p_mat[i, j] <- test$p.value
    }
  }
}

# Build display matrix with stars appended
sig_stars <- function(p) {
  if (is.na(p)) return("")
  if (p < .001) return("***")
  if (p < .01)  return("**")
  if (p < .05)  return("*")
  if (p < .10)  return(".")
  return("")
}

display_mat <- matrix("", n_vars, n_vars, dimnames = list(cor_vars, cor_vars))
for (i in seq_len(n_vars)) {
  for (j in seq_len(n_vars)) {
    if (i == j) {
      display_mat[i, j] <- "—"
    } else {
      display_mat[i, j] <- paste0(r_mat[i, j], sig_stars(p_mat[i, j]))
    }
  }
}

as.data.frame(display_mat) %>%
  kable(caption = "Correlations between compliance and baseline characteristics. * p < .05, ** p < .01, *** p < .001, . p < .10") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Validity Audit

`r committee_note("The committee requested the number of available cases for each Aim.")`

```{r validity-audit}
# Add valid column to full dataset first
study_sample_v <- study_sample %>%
  mutate(valid = check_validity(.))

# Now summarise per participant per part
validity_audit <- study_sample_v %>%
  group_by(participant_id, study_part) %>%
  summarise(
    n_total     = n(),
    n_mood_ok   = sum(!is.na(mood)),
    n_before_ok = sum(!is.na(before_survey)),
    n_smoke_ok  = sum(!is.na(smoke)),
    n_branch_ok = sum(!is.na(pleasant) | !is.na(unpleasant)),
    n_valid     = sum(valid),
    pct_valid   = round(sum(valid) / n() * 100, 1),
    .groups = "drop"
  ) %>%
  mutate(
    study_part = factor(study_part,
                        levels = c(2, 6),
                        labels = c("Part 2", "Part 6")),
    branch_limiting = n_mood_ok == n_smoke_ok &
                      n_smoke_ok == n_before_ok &
                      n_branch_ok < n_mood_ok
  )

# First table: participants where branch is limiting factor
validity_audit %>%
  filter(branch_limiting) %>%
  select(participant_id, study_part, n_total, n_mood_ok,
         n_branch_ok, n_valid, pct_valid) %>%
  arrange(participant_id, study_part) %>%
  kable(caption = "Participants where branch variable (pleasant/unpleasant) is the limiting validity factor") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# Wide format: drop branch_limiting before pivoting
validity_wide <- validity_audit %>%
  select(-branch_limiting) %>%
  pivot_wider(
    id_cols     = participant_id,
    names_from  = study_part,
    values_from = c(n_total, n_mood_ok, n_before_ok,
                    n_smoke_ok, n_branch_ok, n_valid, pct_valid),
    names_glue  = "{study_part}_{.value}"
  )

# Reattach branch_limiting flags per part after widening
branch_flags <- validity_audit %>%
  select(participant_id, study_part, branch_limiting) %>%
  pivot_wider(
    id_cols     = participant_id,
    names_from  = study_part,
    values_from = branch_limiting,
    names_glue  = "{study_part}_branch_limiting"
  )

validity_wide <- validity_wide %>%
  left_join(branch_flags, by = "participant_id")

# Full audit table
validity_wide %>%
  arrange(participant_id) %>%
  kable(caption = "Participant-level validity audit by study part. n_branch_ok = observations with at least one non-missing pleasant/unpleasant value.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  scroll_box(height = "400px")
```

`r section_summary("Available cases for each Aim are determined by study part completion and observation-level validity criteria (see Validity Audit above for full details). Participant 223 is excluded from Part 6 analyses due to zero valid observations during that period. For Aims 1 and 2, which use Part 2 data, the full analytic sample is available. For Aim 3, which uses Part 6 data, the sample is limited to Part 6 completers with valid observations. Counts at varying validity thresholds are reported below. See Section 3 for the CONSORT diagram documenting participant flow through the study.")`

```{r compliance-cutoffs}
# Available cases at different validity thresholds
bind_rows(lapply(c(40, 50, 60), function(cutoff) {
  validity_audit %>%
    group_by(study_part) %>%
    summarise(
      threshold = paste0(cutoff, "%"),
      n_retain  = sum(pct_valid >= cutoff, na.rm = TRUE),
      .groups = "drop"
    )
})) %>%
  pivot_wider(
    names_from  = study_part,
    values_from = n_retain,
    names_glue  = "{study_part}_n_retain"
  ) %>%
  arrange(threshold) %>%
  kable(caption = "Number of participants meeting validity thresholds by study part") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Missingness

```{r missingness}
vis_miss(study_sample %>% select(mood, smoke, before_survey,
                                  pleasant, unpleasant, prior_mood,
                                  smoked_since_last_survey)) +
  labs(title = "Missingness in key EMA variables") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Beep Timing

`r button(text = "Supplementary")`

```{r beep-timing}
study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1) %>%
  mutate(time_of_day = hms::as_hms(init_time),
         weekday = ifelse(wday(init_time, week_start = 1) %in% c(6, 7),
                          "weekend", "weekday")) %>%
  ggplot(aes(x = time_of_day)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  scale_x_time(breaks = scales::date_breaks("2 hours")) +
  facet_grid(weekday ~ .) +
  labs(
    title = "Time of day beeps were started (valid observations only)",
    x     = "Time of day",
    y     = "Number of beeps"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

`r endbutton()`

`r section_summary("Overall compliance was generally good across participants, though variability was present. Compliance trends were examined in relation to baseline participant characteristics to assess whether non-compliance was systematic. A participant-level validity audit identified cases where the branch variable (pleasant/unpleasant) was the limiting factor for observation-level validity — notably, participant 223 had zero valid observations during Part 6 and only 2 during Part 2, rendering them unusable for analyses requiring the full validity criteria. Missingness was concentrated in the branch variables, consistent with the branching structure of the survey. Beep timing followed the expected 2-hour interval sampling scheme across weekdays and weekends.")`

---

# Participant Characteristics and Study Outcomes

`r committee_note("The committee requested examination of how baseline participant characteristics relate to study engagement and key EMA outcomes. Results are organized thematically and are intended to inform covariate selection and interpretation of individual differences in the main analyses.")`

This section examines associations between baseline participant characteristics and two sets of outcomes: (1) study engagement indicators (compliance, attrition, valid observations) and (2) person-level means of key EMA outcomes (craving, mood, pleasant event appraisal ratings, unpleasant event appraisal ratings, and prior mood). All EMA outcome means are computed at the person level from valid observations. For continuous predictors, Pearson correlations with significance stars are reported. For categorical predictors (gender), group means are compared.

```{r characteristics-prep}
# Person-level EMA outcome means (valid observations only)
ema_person_means <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1) %>%
  group_by(participant_id) %>%
  summarise(
    mean_craving    = mean(smoke, na.rm = TRUE),
    mean_mood       = mean(mood, na.rm = TRUE),
    mean_pleasant   = mean(pleasant, na.rm = TRUE),
    mean_unpleasant = mean(unpleasant, na.rm = TRUE),
    mean_prior_mood = mean(prior_mood, na.rm = TRUE),
    n_valid         = sum(!is.na(mood)),
    .groups = "drop"
  )

# Compliance and attrition
engagement <- compliance_df %>%
  select(participant_id, compliance, completion_group) %>%
  mutate(attrition = ifelse(completion_group == "Part 2 Only (3 days)", 1, 0))

# One row per participant with all characteristics
chars_outcomes <- baseline_chars %>%
  select(participant_id, age, gender, total_household_income, bank_balance,
         ftnd_sum, cigsperday, shaps_sim_sum, panas_pos, panas_neg,
         q23, q24, q25, q71, q80) %>%
  left_join(engagement, by = "participant_id") %>%
  left_join(ema_person_means, by = "participant_id")

# Reusable correlation table function with significance stars
cor_table <- function(data, predictors, outcomes,
                      pred_labels, outcome_labels, caption) {
  
  sig_stars <- function(p) {
    if (is.na(p)) return("")
    if (p < .001) return("***")
    if (p < .01)  return("**")
    if (p < .05)  return("*")
    if (p < .10)  return(".")
    return("")
  }
  
  result <- map_dfr(predictors, function(pred) {
    row <- tibble(Predictor = pred_labels[pred])
    for (i in seq_along(outcomes)) {
      out <- outcomes[i]
      test <- cor.test(data[[pred]], data[[out]], use = "pairwise.complete.obs")
      row[[outcome_labels[i]]] <- paste0(round(test$estimate, 2),
                                          sig_stars(test$p.value))
    }
    row
  })
  
  result %>%
    kable(caption = caption) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE)
}

# Variable label lookups
engagement_vars   <- c("compliance", "attrition", "n_valid")
engagement_labels <- c("Compliance", "Attrition\n(1=Part2 only)", "N Valid Obs")

outcome_vars   <- c("mean_craving", "mean_mood", "mean_pleasant",
                    "mean_unpleasant", "mean_prior_mood")
outcome_labels <- c("Craving", "Mood", "Pleasant\nRating",
                    "Unpleasant\nRating", "Prior Mood")
```

## Theme 1: Substance Use History & Developmental Factors

`r committee_note("The committee requested examination of how smoking history and substance use characteristics relate to study engagement and key EMA outcomes.")`

### Association with Study Engagement

```{r theme1-engagement}
pred1 <- c("ftnd_sum", "cigsperday", "q23", "q24", "q25", "q71", "q80")
pred1_labels <- c(
  ftnd_sum    = "Nicotine dependence (FTND)",
  cigsperday  = "Cigarettes per day",
  q23         = "Age at first puff",
  q24         = "Age at first full cigarette",
  q25         = "Age at regular smoking onset",
  q71         = "E-cigarette use (past 30 days)",
  q80         = "Cannabis use (past 30 days)"
)

cor_table(
  data           = chars_outcomes,
  predictors     = pred1,
  outcomes       = engagement_vars,
  pred_labels    = pred1_labels,
  outcome_labels = engagement_labels,
  caption        = "Theme 1: Substance use history and developmental factors — associations with study engagement. * p < .05, ** p < .01, *** p < .001, . p < .10"
)
```

### Association with Key EMA Outcomes

```{r theme1-outcomes}
cor_table(
  data           = chars_outcomes,
  predictors     = pred1,
  outcomes       = outcome_vars,
  pred_labels    = pred1_labels,
  outcome_labels = outcome_labels,
  caption        = "Theme 1: Substance use history and developmental factors — associations with key EMA outcomes. * p < .05, ** p < .01, *** p < .001, . p < .10"
)
```

## Theme 2: Trait Affect & Reward Sensitivity

`r committee_note("The committee requested examination of how trait affect and reward sensitivity relate to study engagement and key EMA outcomes.")`

### Association with Study Engagement

```{r theme2-engagement}
pred2 <- c("panas_pos", "panas_neg", "shaps_sim_sum")
pred2_labels <- c(
  panas_pos    = "Trait positive affect (PANAS)",
  panas_neg    = "Trait negative affect (PANAS)",
  shaps_sim_sum = "Anhedonia (SHAPS)"
)

cor_table(
  data           = chars_outcomes,
  predictors     = pred2,
  outcomes       = engagement_vars,
  pred_labels    = pred2_labels,
  outcome_labels = engagement_labels,
  caption        = "Theme 2: Trait affect and reward sensitivity — associations with study engagement. * p < .05, ** p < .01, *** p < .001, . p < .10"
)
```

### Association with Key EMA Outcomes

```{r theme2-outcomes}
cor_table(
  data           = chars_outcomes,
  predictors     = pred2,
  outcomes       = outcome_vars,
  pred_labels    = pred2_labels,
  outcome_labels = outcome_labels,
  caption        = "Theme 2: Trait affect and reward sensitivity — associations with key EMA outcomes. * p < .05, ** p < .01, *** p < .001, . p < .10"
)
```

## Theme 3: Demographics

`r committee_note("The committee requested examination of how demographic characteristics relate to study engagement and key EMA outcomes.")`

### Continuous Demographics — Association with Study Engagement

```{r theme3-engagement-cont}
pred3_cont <- c("age", "bank_balance")
pred3_cont_labels <- c(
  age          = "Age",
  bank_balance = "Bank balance ($)"
)

cor_table(
  data           = chars_outcomes,
  predictors     = pred3_cont,
  outcomes       = engagement_vars,
  pred_labels    = pred3_cont_labels,
  outcome_labels = engagement_labels,
  caption        = "Theme 3: Continuous demographics — associations with study engagement. * p < .05, ** p < .01, *** p < .001, . p < .10"
)
```

### Continuous Demographics — Association with Key EMA Outcomes

```{r theme3-outcomes-cont}
cor_table(
  data           = chars_outcomes,
  predictors     = pred3_cont,
  outcomes       = outcome_vars,
  pred_labels    = pred3_cont_labels,
  outcome_labels = outcome_labels,
  caption        = "Theme 3: Continuous demographics — associations with key EMA outcomes. * p < .05, ** p < .01, *** p < .001, . p < .10"
)
```

### Gender — Group Means by Study Engagement and EMA Outcomes

```{r theme3-gender}
chars_outcomes %>%
  filter(!is.na(gender)) %>%
  group_by(gender) %>%
  summarise(
    n                = n(),
    mean_compliance  = round(mean(compliance, na.rm = TRUE), 2),
    mean_n_valid     = round(mean(n_valid, na.rm = TRUE), 1),
    pct_attrition    = round(mean(attrition, na.rm = TRUE) * 100, 1),
    mean_craving     = round(mean(mean_craving, na.rm = TRUE), 2),
    mean_mood        = round(mean(mean_mood, na.rm = TRUE), 2),
    mean_pleasant    = round(mean(mean_pleasant, na.rm = TRUE), 2),
    mean_unpleasant  = round(mean(mean_unpleasant, na.rm = TRUE), 2),
    mean_prior_mood  = round(mean(mean_prior_mood, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  rename(
    Gender              = gender,
    N                   = n,
    `Mean Compliance`   = mean_compliance,
    `Mean N Valid`      = mean_n_valid,
    `% Attrition`       = pct_attrition,
    `Mean Craving`      = mean_craving,
    `Mean Mood`         = mean_mood,
    `Mean Pleasant`     = mean_pleasant,
    `Mean Unpleasant`   = mean_unpleasant,
    `Mean Prior Mood`   = mean_prior_mood
  ) %>%
  kable(caption = "Theme 3: Gender — group means for study engagement and key EMA outcomes") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  scroll_box(width = "100%")
```

### Income — Group Means by Study Engagement and EMA Outcomes

```{r theme3-income}
chars_outcomes %>%
  filter(!is.na(total_household_income)) %>%
  mutate(income = as.character(total_household_income)) %>%
  group_by(income) %>%
  summarise(
    n                = n(),
    mean_compliance  = round(mean(compliance, na.rm = TRUE), 2),
    pct_attrition    = round(mean(attrition, na.rm = TRUE) * 100, 1),
    mean_craving     = round(mean(mean_craving, na.rm = TRUE), 2),
    mean_mood        = round(mean(mean_mood, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(income) %>%
  rename(
    `Household Income`  = income,
    N                   = n,
    `Mean Compliance`   = mean_compliance,
    `% Attrition`       = pct_attrition,
    `Mean Craving`      = mean_craving,
    `Mean Mood`         = mean_mood
  ) %>%
  kable(caption = "Theme 3: Household income — group means for study engagement and key EMA outcomes. All income categories shown.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

`r section_summary("Associations between baseline participant characteristics and study engagement and EMA outcomes are examined across three thematic groupings: substance use history and developmental factors, trait affect and reward sensitivity, and demographics. Results are descriptive and intended to inform covariate selection, interpretation of individual differences, and potential confounds in the main analyses.")`

---

# EMA Item Descriptives

This section describes the distributions, variance structure, and temporal patterns of the key EMA variables.

## Distributions

`r button(text = "Description")`

```{r ema-distributions}
ema_vars <- c("mood", "smoke", "pleasant", "unpleasant", "prior_mood")

study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1) %>%
  select(all_of(ema_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  labs(
    title = "Distributions of key EMA variables (valid observations)",
    x     = "Value",
    y     = "Count"
  ) +
  theme_minimal()
```

`r endbutton()`

## Intraclass Correlations (ICCs)

`r committee_note("The committee requested ICCs for key EMA variables to distinguish between- and within-person variance and justify modeling decisions.")`

```{r iccs}
# Fit null models (random intercept only) and extract ICC
# ICC = between-person variance / (between-person variance + within-person variance)

compute_icc_lme4 <- function(df, var, id = "participant_id") {
  
  df_clean <- df %>%
    filter(!is.na(.data[[var]])) %>%
    rename(y = all_of(var), id_var = all_of(id))
  
  # Fit null model: outcome ~ 1 + (1 | participant)
  model <- lmer(y ~ 1 + (1 | id_var), data = df_clean, REML = TRUE)
  
  vc <- as.data.frame(VarCorr(model))
  
  between_var <- vc$vcov[vc$grp == "id_var"]
  within_var  <- vc$vcov[vc$grp == "Residual"]
  total_var   <- between_var + within_var
  icc         <- between_var / total_var
  
  tibble(
    variable    = var,
    icc         = round(icc, 3),
    between_var = round(between_var, 3),
    within_var  = round(within_var, 3),
    total_var   = round(total_var, 3),
    pct_between = round(icc * 100, 1),
    pct_within  = round((1 - icc) * 100, 1)
  )
}

ema_valid <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1)

icc_results <- bind_rows(
  lapply(c("mood", "smoke", "pleasant", "unpleasant", "prior_mood"),
         compute_icc_lme4, df = ema_valid)
)

icc_results %>%
  kable(caption = "Intraclass correlations (ICCs) for key EMA variables. ICCs were estimated using null models (random intercept only) fit with REML via lme4. Between-person variance reflects stable individual differences; within-person variance reflects moment-to-moment fluctuation.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Temporal Patterns

`r button(text = "Description")`

```{r temporal-beepno}
# Means by beep number within day
study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1) %>%
  select(beepno, all_of(c("mood", "smoke", "pleasant", "unpleasant", "prior_mood"))) %>%
  pivot_longer(-beepno, names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(beepno, variable) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = beepno, y = mean_value)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Mean EMA variable values by beep number within day",
    x     = "Beep number",
    y     = "Mean value"
  ) +
  theme_minimal()
```

```{r temporal-dayno}
# Means by study day
study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1) %>%
  select(dayno, all_of(c("mood", "smoke", "pleasant", "unpleasant", "prior_mood"))) %>%
  pivot_longer(-dayno, names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(dayno, variable) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = dayno, y = mean_value)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 3.5, linetype = "dashed", color = "gray50") +
  annotate("text", x = 2, y = Inf, label = "Part 2", vjust = 1.5,
           size = 3, color = "gray40") +
  annotate("text", x = 7, y = Inf, label = "Part 6", vjust = 1.5,
           size = 3, color = "gray40") +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Mean EMA variable values by study day",
    x     = "Study day",
    y     = "Mean value"
  ) +
  theme_minimal()
```

`r endbutton()`

## Condition-Level Means

`r committee_note("The committee requested means for mood and craving across unpleasant, pleasant, and neutral event appraisal conditions.")`

```{r condition-means}
study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1, !is.na(before_survey)) %>%
  group_by(before_survey) %>%
  summarise(
    n          = n(),
    mean_mood  = round(mean(mood, na.rm = TRUE), 2),
    sd_mood    = round(sd(mood, na.rm = TRUE), 2),
    mean_smoke = round(mean(smoke, na.rm = TRUE), 2),
    sd_smoke   = round(sd(smoke, na.rm = TRUE), 2)
  ) %>%
  kable(caption = "Means and SDs for mood and craving by event appraisal condition") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

`r section_summary("Distributions of key EMA variables (mood, craving, pleasant, unpleasant, prior mood) are described above. ICCs from null models indicate the proportion of variance attributable to stable between-person differences versus moment-to-moment within-person fluctuation — results inform whether multilevel modeling is warranted and how predictors should be centered. Temporal patterns by beep number and study day are examined to identify potential time-of-day and study-phase effects. Condition-level means show how mood and craving differ across pleasant, unpleasant, and neutral event appraisal contexts, providing a preliminary look at the central contrasts for Aims 1 and 2.")`

---

# Smoking Behavior

`r committee_note("This section addresses several committee questions about cigarette use behavior over the course of the study.")`

## Person-Level Cigarette Use

`r committee_note("The committee requested a description of person-level cigarette use behavior over the course of the study.")`

`r button(text = "Description")`

```{r person-smoking}
# Daily smoking totals per participant per day
smoking_person <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1, !is.na(smoked_since_last_survey), !is.na(dayno)) %>%
  group_by(participant_id, dayno) %>%
  summarise(
    daily_smoked = sum(smoked_since_last_survey, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(completer = ifelse(participant_id %in% part6_completers,
                            "Part 6 Completer", "Part 2 Only"))

# --- Plot 1: Color-coded spaghetti ---
ggplot(smoking_person, aes(x = dayno, y = daily_smoked,
                            group = participant_id,
                            color = completer,
                            alpha = completer)) +
  geom_line() +
  geom_vline(xintercept = 3.5, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("Part 6 Completer" = "steelblue",
                                 "Part 2 Only"      = "tomato")) +
  scale_alpha_manual(values = c("Part 6 Completer" = 0.5,
                                 "Part 2 Only"      = 0.3)) +
  scale_x_continuous(breaks = 1:10) +
  annotate("text", x = 2, y = Inf, label = "Part 2\n(Baseline)",
           vjust = 1.5, size = 3, color = "gray40") +
  annotate("text", x = 7, y = Inf, label = "Part 6\n(Incentive)",
           vjust = 1.5, size = 3, color = "gray40") +
  labs(
    title = "Person-level daily cigarette use over the study",
    x     = "Study day",
    y     = "Cigarettes reported",
    color = "Completion status",
    alpha = "Completion status"
  ) +
  theme_minimal()

# --- Plot 2: Part 6 completers, separate smoothers by study part ---
smoking_completers <- smoking_person %>%
  filter(completer == "Part 6 Completer") %>%
  mutate(part = ifelse(dayno <= 3, "Part 2 (Baseline)", "Part 6 (Incentive)"))

ggplot(smoking_completers, aes(x = dayno, y = daily_smoked)) +
  geom_line(aes(group = participant_id), color = "steelblue", alpha = 0.3) +
  geom_smooth(aes(group = part, color = part),
              method = "lm", se = TRUE, linewidth = 1) +
  geom_vline(xintercept = 3.5, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("Part 2 (Baseline)"  = "gray40",
                                 "Part 6 (Incentive)" = "darkblue")) +
  scale_x_continuous(breaks = 1:10) +
  annotate("text", x = 2, y = Inf, label = "Part 2\n(Baseline)",
           vjust = 1.5, size = 3, color = "gray40") +
  annotate("text", x = 7, y = Inf, label = "Part 6\n(Incentive)",
           vjust = 1.5, size = 3, color = "gray40") +
  labs(
    title = "Daily cigarette use among Part 6 completers",
    subtitle = "Individual trajectories (light) with separate smoothers by study part",
    x     = "Study day",
    y     = "Cigarettes reported",
    color = "Study part"
  ) +
  theme_minimal()

# --- Plot 3: Deviation from baseline, color-coded by change category ---

# Compute Part 2 baseline mean per person
baseline_ref <- smoking_completers %>%
  filter(dayno <= 3) %>%
  group_by(participant_id) %>%
  summarise(baseline_mean = mean(daily_smoked, na.rm = TRUE),
            .groups = "drop")

# Part 6 data only, with deviation from baseline
part6_deviation <- smoking_completers %>%
  filter(dayno > 3) %>%
  left_join(baseline_ref, by = "participant_id") %>%
  mutate(deviation = daily_smoked - baseline_mean)

# Classify each participant by their mean Part 6 deviation
# Quitter  = Part 6 mean == 0 (or effectively zero, <= 0.1 to handle rounding)
# Reducer  = Part 6 mean < baseline by > 20%
# Stable   = within 20% of baseline in either direction
# Increaser = Part 6 mean > baseline by > 20%
person_class <- part6_deviation %>%
  group_by(participant_id, baseline_mean) %>%
  summarise(part6_mean = mean(daily_smoked, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(
    change_cat = case_when(
      part6_mean <= 0.1                                    ~ "Quitter",
      part6_mean < baseline_mean * 0.80                   ~ "Reducer",
      part6_mean > baseline_mean * 1.20                   ~ "Increaser",
      TRUE                                                 ~ "Stable"
    )
  )

# Print classification counts — useful to verify your hunch
cat("Change category counts:\n")
print(table(person_class$change_cat))

# Merge classification back
part6_deviation <- part6_deviation %>%
  left_join(person_class %>% select(participant_id, change_cat),
            by = "participant_id")

# Plot
ggplot(part6_deviation, aes(x = dayno, y = deviation,
                             group = participant_id,
                             color = change_cat)) +
  geom_line(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  scale_color_manual(
    values = c(
      "Quitter"   = "darkgreen",
      "Reducer"   = "steelblue",
      "Stable"    = "gray60",
      "Increaser" = "tomato"
    )
  ) +
  scale_x_continuous(breaks = 4:10) +
  annotate("text", x = 6.5, y = Inf,
           label = "Dashed line = individual Part 2 baseline mean",
           vjust = 1.5, size = 2.8, color = "gray40") +
  labs(
    title = "Part 6 smoking deviations from individual Part 2 baselines",
    subtitle = "Quitter = zero Part 6 smoking; Reducer = >20% below baseline;\nStable = within 20%; Increaser = >20% above baseline",
    x     = "Study day",
    y     = "Deviation from baseline (cigarettes reported)",
    color = "Change category"
  ) +
  theme_minimal()
```

`r endbutton()`

## 3-Day Baseline Smoking

`r committee_note("The committee requested that 3-day EMA cigarette smoking data be used to calculate each participant's typical smoking behavior as a baseline metric for assessing relative deviations during Part 6.")`

```{r baseline-smoking}
# --- Table 1: Part 2 vs Part 6 smoking summary for completers ---

# Part 2 summary — all participants
part2_all <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1, study_part == 2, !is.na(smoked_since_last_survey)) %>%
  group_by(participant_id) %>%
  summarise(
    baseline_total = sum(smoked_since_last_survey, na.rm = TRUE),
    baseline_days  = n_distinct(dayno),
    baseline_daily = round(baseline_total / baseline_days, 2),
    .groups = "drop"
  )

# Part 2 summary — completers only
part2_completers <- part2_all %>%
  filter(participant_id %in% part6_completers)

# Part 6 summary — completers only
part6_summary <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1, study_part == 6, !is.na(smoked_since_last_survey)) %>%
  group_by(participant_id) %>%
  summarise(
    part6_total = sum(smoked_since_last_survey, na.rm = TRUE),
    part6_days  = n_distinct(dayno),
    part6_daily = round(part6_total / part6_days, 2),
    .groups = "drop"
  )

# Build comparison table
tibble(
  Sample = c("Part 2 completers",
           "Part 6 completers — Part 2 (baseline)",
           "Part 6 completers — Part 6 (incentive)"),
N = c(
  nrow(part2_completers),
  nrow(part2_completers),
  nrow(part6_summary)
),
Mean_daily = c(
  round(mean(part2_completers$baseline_daily, na.rm = TRUE), 2),
  round(mean(part2_completers$baseline_daily, na.rm = TRUE), 2),
  round(mean(part6_summary$part6_daily, na.rm = TRUE), 2)
),
  SD_daily = c(
    round(sd(part2_completers$baseline_daily, na.rm = TRUE), 2),
    round(sd(part2_completers$baseline_daily, na.rm = TRUE), 2),
    round(sd(part6_summary$part6_daily, na.rm = TRUE), 2)
  ),
  Min_daily = c(
    min(part2_completers$baseline_daily, na.rm = TRUE),
    min(part2_completers$baseline_daily, na.rm = TRUE),
    min(part6_summary$part6_daily, na.rm = TRUE)
  ),
  Max_daily = c(
    max(part2_completers$baseline_daily, na.rm = TRUE),
    max(part2_completers$baseline_daily, na.rm = TRUE),
    max(part6_summary$part6_daily, na.rm = TRUE)
  )
) %>%
  kable(caption = "Daily cigarette use summary by sample and study part") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# --- Table 2: Change category summary ---
# Recompute person_class here in case chunk is run independently
person_class_summary <- part2_completers %>%
  left_join(part6_summary, by = "participant_id") %>%
  mutate(
    change_cat = case_when(
      part6_daily <= 0.1                          ~ "Quitter",
      part6_daily < baseline_daily * 0.80         ~ "Reducer",
      part6_daily > baseline_daily * 1.20         ~ "Increaser",
      TRUE                                         ~ "Stable"
    )
  )

person_class_summary %>%
  group_by(change_cat) %>%
  summarise(
    n                    = n(),
    pct                  = round(n() / nrow(person_class_summary) * 100, 1),
    mean_baseline_daily  = round(mean(baseline_daily, na.rm = TRUE), 2),
    sd_baseline_daily    = round(sd(baseline_daily, na.rm = TRUE), 2),
    mean_part6_daily     = round(mean(part6_daily, na.rm = TRUE), 2),
    sd_part6_daily       = round(sd(part6_daily, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(n)) %>%
  kable(caption = "Part 6 smoking change categories relative to Part 2 baseline. Quitter = mean Part 6 daily smoking <= 0.1; Reducer = >20% below baseline; Increaser = >20% above baseline; Stable = within 20% of baseline.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

```
> **Note:** One participant (ID 223) is excluded from Part 6 smoking summaries due to missing branch variable data (pleasant/unpleasant) across all Part 6 observations, resulting in zero valid observations for that participant during Part 6.

`r section_summary("Person-level cigarette use trajectories are characterized across the full study period, with Part 6 completers distinguished from Part 2-only participants. Among Part 6 completers, within-day smoking patterns (by beep number) and time-of-day effects are examined separately for Part 2 and Part 6. Participants are classified into change categories (Quitter, Reducer, Stable, Increaser) based on their mean Part 6 smoking relative to their individual Part 2 baseline. CO-EMA agreement is examined to assess the validity of self-reported smoking behavior against biochemical verification.")`

## CO and EMA Alignment

`r committee_note("The committee asked whether the CO data align with the EMA data on cigarette smoking for each participant.")`

`r button(text = "Description")`

```{r co-ema-alignment}
# Compare CO-verified smoking status with EMA-reported smoking
co_ema <- study_sample %>%
  filter(!is.na(co_level), !is.na(co_state_smoked)) %>%
  group_by(participant_id, dayno) %>%
  summarise(
    ema_smoked_any  = any(smoked_since_last_survey > 0, na.rm = TRUE),
    co_smoked       = first(co_state_smoked),
    co_level        = first(co_level),
    .groups = "drop"
  ) %>%
  filter(!is.na(co_smoked))

# Agreement table
co_ema %>%
  mutate(
    agreement = case_when(
      ema_smoked_any == TRUE  & co_smoked == TRUE  ~ "Both: smoked",
      ema_smoked_any == FALSE & co_smoked == FALSE ~ "Both: did not smoke",
      ema_smoked_any == TRUE  & co_smoked == FALSE ~ "EMA: smoked, CO: did not",
      ema_smoked_any == FALSE & co_smoked == TRUE  ~ "EMA: did not, CO: smoked",
      TRUE ~ "Indeterminate"
    )
  ) %>%
  count(agreement) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  kable(caption = "Agreement between EMA-reported and CO-verified smoking") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

`r endbutton()`

---

# Abstinence Incentive

`r committee_note("This section addresses the committee's question about the effect of the abstinence incentive. Analyses are framed descriptively to inform future modeling decisions.")`

```{r incentive-effect}
# Compare smoking behavior between Part 2 and Part 6
incentive_compare <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1, !is.na(smoked_since_last_survey)) %>%
  group_by(participant_id, study_part) %>%
  summarise(
    mean_smoked = mean(smoked_since_last_survey, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(study_part = factor(study_part,
                              levels = c(2, 6),
                              labels = c("Part 2 (Baseline)", "Part 6 (Incentive)")))

# Summary
incentive_compare %>%
  group_by(study_part) %>%
  summarise(
    n           = n(),
    mean_smoked = round(mean(mean_smoked, na.rm = TRUE), 2),
    sd_smoked   = round(sd(mean_smoked, na.rm = TRUE), 2)
  ) %>%
  kable(caption = "Mean cigarettes reported per beep by study part") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r smoking-momentary}

# --- Plot 1: Mean cigarettes by beep number, Part 2 only, full sample ---
plot1_data <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1, study_part == 2, !is.na(smoked_since_last_survey),
         !is.na(beepno)) %>%
  group_by(beepno) %>%
  summarise(
    mean_smoked = mean(smoked_since_last_survey, na.rm = TRUE),
    sd_smoked   = sd(smoked_since_last_survey, na.rm = TRUE),
    n           = n(),
    se_smoked   = sd_smoked / sqrt(n),
    .groups = "drop"
  )

ggplot(plot1_data, aes(x = beepno, y = mean_smoked)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_errorbar(aes(ymin = mean_smoked - se_smoked,
                    ymax = mean_smoked + se_smoked),
                width = 0.2, color = "steelblue", alpha = 0.6) +
  scale_x_continuous(breaks = 1:6) +
  labs(
    title    = "Mean cigarettes reported by beep number (Part 2, full sample)",
    subtitle = "Error bars = SE",
    x        = "Beep number",
    y        = "Mean cigarettes reported"
  ) +
  theme_minimal()

# --- Plot 2: Mean cigarettes by beep number, Part 6 completers,
#             Part 2 vs Part 6 ---
plot2_data <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1,
         participant_id %in% valid_completers,
         !is.na(smoked_since_last_survey),
         !is.na(beepno)) %>%
  mutate(study_part = factor(study_part,
                              levels = c(2, 6),
                              labels = c("Part 2 (Baseline)",
                                         "Part 6 (Incentive)"))) %>%
  group_by(beepno, study_part) %>%
  summarise(
    mean_smoked = mean(smoked_since_last_survey, na.rm = TRUE),
    sd_smoked   = sd(smoked_since_last_survey, na.rm = TRUE),
    n           = n(),
    se_smoked   = sd_smoked / sqrt(n),
    .groups = "drop"
  )

ggplot(plot2_data, aes(x = beepno, y = mean_smoked,
                        color = study_part, group = study_part)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_smoked - se_smoked,
                    ymax = mean_smoked + se_smoked),
                width = 0.2, alpha = 0.6) +
  scale_color_manual(values = c("Part 2 (Baseline)"  = "gray40",
                                 "Part 6 (Incentive)" = "darkblue")) +
  scale_x_continuous(breaks = 1:6) +
  labs(
    title    = "Mean cigarettes reported by beep number (Part 6 completers)",
    subtitle = "Error bars = SE",
    x        = "Beep number",
    y        = "Mean cigarettes reported",
    color    = "Study part"
  ) +
  theme_minimal()

# --- Plot 3: Time of day effects by study part, Part 6 completers ---
plot3_data <- study_sample %>%
  mutate(valid = check_validity(.)) %>%
  filter(valid == 1,
         participant_id %in% valid_completers,
         !is.na(smoked_since_last_survey),
         !is.na(completion_time)) %>%
  mutate(
    hour       = hour(completion_time),
    study_part = factor(study_part,
                         levels = c(2, 6),
                         labels = c("Part 2 (Baseline)",
                                    "Part 6 (Incentive)"))
  ) %>%
  group_by(hour, study_part) %>%
  summarise(
    mean_smoked = mean(smoked_since_last_survey, na.rm = TRUE),
    sd_smoked   = sd(smoked_since_last_survey, na.rm = TRUE),
    n           = n(),
    se_smoked   = sd_smoked / sqrt(n),
    .groups = "drop"
  )

ggplot(plot3_data, aes(x = hour, y = mean_smoked,
                        color = study_part, group = study_part)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_smoked - se_smoked,
                    ymax = mean_smoked + se_smoked),
                width = 0.3, alpha = 0.6) +
  scale_color_manual(values = c("Part 2 (Baseline)"  = "gray40",
                                 "Part 6 (Incentive)" = "darkblue")) +
  scale_x_continuous(breaks = seq(6, 22, by = 2)) +
  labs(
    title    = "Mean cigarettes reported by time of day (Part 6 completers)",
    subtitle = "Error bars = SE",
    x        = "Hour of day (completion time)",
    y        = "Mean cigarettes reported",
    color    = "Study part"
  ) +
  theme_minimal()
```

```{r incentive-plot}
ggplot(incentive_compare, aes(x = study_part, y = mean_smoked,
                               fill = study_part)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1) +
  labs(
    title = "Cigarette use by study part",
    x     = "Study part",
    y     = "Mean cigarettes reported per beep"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

*Note: Formal tests of the incentive effect are reserved for the main analyses. The above is intended to inform whether smoking behavior changed between parts and whether this should be considered in future modeling.*

`r section_summary("Descriptive comparisons of smoking behavior between Part 2 and Part 6 suggest a meaningful reduction in cigarettes reported per beep during the incentivized abstinence period. Individual variability in response to the incentive is substantial, as reflected in the wide distributions. These results are framed descriptively and will inform whether study part or incentive-related variables need to be accounted for in the main analyses.")`

---

# Available Cases


`r committee_note("The committee requested the number of available cases for each Aim, including a CONSORT diagram.")`

Available cases by Aim are summarized in the Validity Audit section above. See Section 3 for the CONSORT diagram.

---

# References

Revol, J., Carlier, C., Lafit, G., Verhees, M., Sels, L., & Ceulemans, E. (2024). Preprocessing ESM data: A step-by-step framework, tutorial website, R package, and reporting templates. *Advances in Methods and Practices in Psychological Science, 7*(4), 1--18. https://doi.org/10.1177/25152459241256609

Siepe, B. S., Rieble, C., Tutunji, R., Rimpler, A., März, J., Proppert, R. K. K., & Fried, E. I. (2024). Understanding EMA data: A tutorial on exploring item performance in ecological momentary assessment data. *PsyArXiv*. https://doi.org/10.31234/osf.io/dvj8g
